---
title: "Session_1"
author: "Saewon Park"
date: "November 3, 2018"
output: pdf_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```






## Session 1: Basics ##

R is object oriented. An object can be a value, vector, dataframe, list, plot, and even a function. 

Anything you can save is an object and R works by calling these objects and using them.





```{r}

## An object can be a single value (AKA: variable)


#We assign a value to a variable using "<-"

a <- 4 # 4 is assigned to "a"
a # calling a
a + 5 # since we assigned a numeric value to a, we can perform arithmetric operations on it 

b <- 10
a*b # objects can interact with each other

c <- a*b # we can also assign variables to another variable
c

```







## 2. Vectors

We can also store multiple elements of the same type. This is called a vector.



```{r }

d <- c(1,2,3) #c() is short for concatenate
d



d <- 2:3 #a sequence of number 1 through 3



d*10 #when the elements of a vector are all numbers, we can multiply each element with a number



e <- c("Red", "Orange", "Green") # the values assigned to an object aren't limited to numbers
e




f <- c(d,e)
f

```






### 2.1 Element types in R

There are mainly three types of elements in R for vectors: numeric (including integer), character, and logical

```{r }
#we can check the data type of an object by using the class() function
class(d)
class(e)


#the logical data type consists of TRUE and FALSE, which can be shortened to T and F
g <- c(TRUE, FALSE, T, T, F)
class(g)

```




What do you think happens with vectors that have both numbers and character? Check the class of "f".




```{r }

class(f)
#when numbers and characters are put in one vector, the numbers are coerced as characters

```







## 3. Subsetting



The order of the elements in a vector are stored as an index. We can use indices to extract specific elements from a list.




```{r }

f[4] #extracting the 4th element from f
#note that we use [] for indices and not ()


f[4:6] #extracting the 4th to the 6th elements from f  


f[c(1,5,6)] #extracting the first, 5th, 6th elements from f






# We can use this to reassign values
f[4] <- "Yellow" #assigning the 4th element of f the value "Yellow"
f


```






## 4. Repeating, Sequences and Matrices



### 4.1 Repetition

```{r }

# Our first function: rep()
h <- rep(x ="Some Text", times = 5)
h



?rep # ?+function name/package name will bring up the help page
     # you can always use google for help as well




#we can also repeat a sequence of multiple elements
h <- rep(1:3, 5) #if the arguments are in order, we can remove the "="
h

```







### 4.2 Sequences  



```{r }

i <- seq(from = 30, to = 300, by = 1) # outputs a sequence of numbers from 30 to 300, increasing by 1 every time.
i


j <- seq(50, 200, 5) #50 to 200, increasing by 5 every time
j




```







### 4.3 Matrices - a new data type


Vectors stored data in one-dimension. Matries store data in two-dimensions. Matrices have rows and columns and they all have to be of the same element type.




```{r }

k <- cbind(i, j) #binding vectors i and j as columns

l <- rbind(i, j) #binding vectors i and j as rows

# Note: What do the warnings mean? What happened to the shorter vector in the matrices?

```







## Quick Review


a) create a vector called x1 that contains the numbers 1, 10, and 20.

b) create a vector called x2 that contains every number in the sequence 100, 95, 90, 85 ... -95, -100.

c) create a vector called x3 that contains that first element of x1 and the 6th element of x2.




```{r }

#Answers

x1 <- c(1, 10, 20)

x2 <- seq(100, -100, -5)

x3 <- c(x1[1], x2[6])


```





## 5. Importing Data

Now, let's look at an actual dataset. "fearon03.csv" is a dataset containing information about civil wars.



```{r }

#setwd("/Users/saewonpark/Documents/McGill/DSC/")  #use the path of the directory with "fearon03.csv" in your own computer


civilwar <- read.csv("fearon03.csv") 
#This creates a dataframe, which is like a matrix but with row/column names. There can be different data type for each column


class(civilwar)



# Other packages you can use to open csv: foreign, readxl


```





## 6. Selecting within a dataframe


1. Using column (variable) names



```{r }

#we can select a column using the column name
civilwar$country

#to see all column names
names(civilwar)



#this only only returns the unique values in the "country column"
unique(civilwar$country)


#using unique() we create a vector of all the country names in the dataset
country_names <- unique(civilwar$country)

```


2. Using indexing

```{r }

#like we did with vectors, we can subset dataframes and matrices using indices within square brackets.
#[row_number, column_number]

civilwar[1, ] #first row, all columns


civilwar[ ,2] #all rows, second column


civilwar[1,2] #first row, second column


civilwar[1, 2:4] #what would be the output of this?


```



3. Filtering through the dataset


```{r }

civilwar[civilwar$country == "CANADA", ] #gets all the observations where Canada is the country
# notice that we use double equal sign for comparison (and single equal sign to assign variables)


civilwar[civilwar$country == "CANADA", 8] #gets the 8th column (region) for just Canadian observations


civilwar[civilwar$country == "CANADA", "gdppc"] #same as above, but using column name


#AND, OR, %in% notations
civilwar[civilwar$country == "CANADA" | civilwar$country == "USA", ] # | means OR

civilwar[civilwar$country == "CANADA" & civilwar$country == "USA", ] # & means AND

civilwar[civilwar$country %in% c("CANADA","USA"), ] # %in% finds matches between the vectors coming before and after this special notation


```





Creating new dataframes from subsets


```{r }

# Creating new matrices:


# by row filter:
canada <- civilwar[civilwar$country == "CANADA", ]

can_usa <- civilwar[civilwar$country %in% c("CANADA","USA"), ] 



# by column filter: 
gdppop <- civilwar[ , c("gdppc", "pop")]



```






## Quick Review

a) Create a vector called regions that lists all unique regions in the dataset

b) Subset the civilwar dataset to include only BURUNDI, ITALY, and ALBANIA. Assign this to a new variable called bur_ita_alb





```{r }

#Answers

#a)
regions <- unique(civilwar$region)
#or
regions <- unique(civilwar[ , 8])



#b)
bur_ita_alb <- civilwar[civilwar$country %in% c("BURUNDI", "ITALY", "ALBANIA"), ]
#or
bur_ita_alb <- civilwar[civilwar$country == "BURUNDI" |
                          civilwar$country == "ITALY" |
                          civilwar$country == "ALBANIA", ]

```






## 7. Descriptive Statistics



```{r }

sum(civilwar$war) #sum of all instances of war (where war is coded as 1)

mean(civilwar$war) #mean of "war" column

mean(civilwar$polity2) #mean() does not work when there are NAs



#Other useful functions for descriptives & cleaning data:

is.na(civilwar$polity2) #checks if there are any NAs in this vector

sum(is.na(civilwar$polity2)) #sums up all instances of NAs


mean(civilwar$polity2, na.rm = TRUE) #we can remove the NAs and then take the mean


#minimum and maximum
min(civilwar$war)
max(civilwar$war)

median(civilwar$polity2)

#summary statistics at a glance
summary(civilwar$polity2)
summary(civilwar$region)


#you can use summary() for the entire dataset
summary(civilwar)


#counts the number of observations.
nrow(civilwar)



# Subsetting + Descriptive statistics
mean(civilwar$gdppc[civilwar$country == "CANADA"]) #mean of the gdp per capita for Canadian observations


civilwar$country[civilwar$mtn == max(civilwar$mtn, na.rm = TRUE)] #country with the highest percentage of mountains
# Note that this command evaluates all rows to see if the max() statement is true


unique(civilwar$country[civilwar$mtn == max(civilwar$mtn, na.rm = TRUE)]) #to get just one value as the output


```






## 8. Creating new variables



```{r }

summary(civilwar$pop)


hist(civilwar$pop) #histogram of population (shows distribution)

civilwar$popln <- log(civilwar$pop) #creating a new variable for the log of population

View(civilwar) #opens a tab where you can see the dataframe

hist(civilwar$popln) #histogram of new variable, you can see that the distribution is more normal


#let's create an entirely new variable using polity scores

summary(civilwar$polity2)

civilwar$democracy <- ifelse(civilwar$polity2 > 5, 1, 0) #ifelse function
#creating a dummy variables that defines all countries with a polity score greater than 5 as a democracy

mean(civilwar$democracy, na.rm = TRUE)


```





## 9. Recoding variables 



In most datasets you will work with, there will be issues in the dataset that requires fixing. This stage is called data cleaning.




```{r }

## Problem: Mountainous terrain codes missing data as -99 instead of NA
summary(civilwar$mtn) #this skews the mean


#recode -99 as NA
civilwar$mtn[civilwar$mtn == -99] <- NA


# Or, recode using ifelse
civilwar$mtn2 <- ifelse(civilwar$mtn == -99, NA, civilwar$mtn)


summary(civilwar$mtn)
summary(civilwar$mtn2)
#they are the same



# Sometimes you may find that a numeric variable is coded as character type. To change it back to numeric:

civilwar$pop <- as.character(civilwar$pop) #to create an example, we change the population column to character
class(civilwar$pop)

civilwar$pop <- as.numeric(civilwar$pop) #change back to numeric
class(civilwar$pop)


### factor to numeric: as.numeric(as.character(x))

# Using dplyr's group_by() and mutate() functions -- next week! 



```



## 10. Saving data

```{r }
write.csv(civilwar, "fearon03_v2.csv")
```



## 11. Clearing environment

```{r }
remove(list = ls())
```



