---
title: 'Session 2: Practice Problems'
author: "Andrew McCormack"
date: '2018-11-13'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE,
                      include = TRUE)
library(dplyr)
```

```{r echo = TRUE}
setwd("/Users/andrewmccormack/Documents/DSC/")
civilwar <- read.csv("fearon03.csv")
```

## Problem 1

1. Load the civil war dataset.
2. Select the `pop`, `mtn`, and `polity2` variables. Create a data frame with just those three variables along with the variables `country` and `year`. Do this using the `select()` function from `dplyr`.

```{r }
cw <- civilwar %>%
  select(country, year, pop, mtn, polity2)
```


## Problem 2

1. There is an issue with the mountainous terrain variable, `mtn`. Instead of `NA` values, missing data was coded as `-99`. This won't do. Ask R for a summary of the mountain variable to verify this. 

```{r }
summary(cw$mtn)
```

2. Create a new variable called `mtn2` that is the same as `mtn` except with the `-99` values replaced with `NA`s. Don't alter the original mountain variable.

```{r }
cw$mtn2 <- cw$mtn
cw$mtn2[cw$mtn2 == -99] <- NA
```

3. Bonus: there are two ways to solve problem 2---one using `dplyr` and the other using the basic functions from last week. Try using `dplyr` to solve this problem (hint: `mutate()` and `ifelse()` might come in handy here).

```{r }
cw <- cw %>%
  mutate(mtn2 = ifelse(mtn == -99, NA, mtn))
```

## Problem 3

1. Using the original data, create a ``dummy'' variable for observations where the value is 1 if the country is located in Sub-Saharan Africa and 0 otherwise. 

```{r}
unique(civilwar$region)

cw2 <- civilwar %>%
  mutate(sub_saharan_africa = ifelse(region == "sub-saharan africa", 1,0))
```


2. Select a numeric variable and calculate its average separately for countries in Sub-Saharan Africa and countries elsewhere. Try doing this using the pipe operator in `dplyr`. What is the difference in the two means?

```{r}
# Base method
cw2$gdppc[cw2$sub_saharan_africa == 1] %>%
  mean(na.rm = TRUE)

cw2$gdppc[cw2$sub_saharan_africa == 0] %>%
  mean(na.rm = TRUE)
```
```{r}
# dplyr method
cw2 %>%
  group_by(sub_saharan_africa) %>%
  summarise(gdppc_mean = mean(gdppc, na.rm = TRUE))
```

## Problem 4

1. Filter the original dataset to include only the years 1989 and 1999.

2. Find the means for `polity2` by region and year (hint: you can use two variables in `group_by()`). Did any regions experience democratic backsliding between 1989 and 1999? If so, which ones?

```{r }
civilwar %>%
  filter(year %in% c(1989, 1999)) %>%
  group_by(region, year) %>%
  summarise(democracy_mean = mean(polity2, na.rm = T))
```
*Bonus: complete steps 1 and 2 simultaneously using the pipe operator. 
